Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-09-15 manuscript/manuscript.tex
renamed ref.bib to refs.bib
rewrote recommendation section with principles and actionable steps

\section{NYC 311 Service Request Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
\jy{should we mention the system upgrade just before the pandemic to
  document this historical event?}
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system.

@@

\section{Data Cleansing Issues} 
\label{sec:issues}

@@

\section{Recommendations} 
\label{sec:recommendations}
We propose the following recommendations for the Open Data Team to 
address the data cleanliness issues discussed in this paper.

\begin{itemize}
\item  Undertake a dedicated effort 
    to investigate these data anomalies, confirm the veracity of the 
    issues, and collaborate with the appropriate Agency Open Data 
    Representatives to find resolutions. This effort may be complex 
    and require software changes. Coordination and mutual approval 
    of changes will be key, requiring strong support from senior 
    leadership.

    \item Standards and protocols should be established, agreed upon, 
    and enforced. This includes defining the domain of legal values, 
    ensuring data accuracy, implementing address normalization, 
    clarifying data usage, and providing detailed data definitions.

    \item Resolve many potential data integrity issues
    through enforcement by software. For instance, software should 
    prevent invalid entries like erroneous zip codes. Similarly, fields 
    like \texttt{created\_date} and \texttt{closed\_date}, which 
    measure responsiveness, should have logic checks to prevent 
    illogical sequences, such as an SR being ``closed'' before 
    being ``opened.''

    \item The large spikes in SR creation and closure at midnight and 
    noon likely point to automated processes from external Agency 
    integrations. These distortions should be eliminated in favor of 
    capturing accurate SR durations with precise timestamps.

    \item As discussed, there are six ``computed'' fields that are not 
    listed in the Data Dictionary. Some of these fields contain 
    significant inaccuracies. We recommend updating the Data 
    Dictionary and, if necessary, masking these fields from public use 
    until their accuracy is improved.

    \item Duplicate fields should be reviewed for removal. In some 
    cases, such as \texttt{borough} and \texttt{park\_borough}, the 
    duplication is 100\%. In others, the match is 80-90\%. 
    These near-duplicates should be assessed to determine their 
    relevance in the 311 SR dataset. Initial analysis suggests that 
    removing these fields could reduce file size by up to 12\%, with 
    further reductions possible upon deeper review.

    \item Fields with very low data presence, such as 
    \texttt{taxi\_company\_borough}, populated only 0.06\% of the time, 
    should be evaluated to determine if they are necessary.

    \item The Data Dictionary, last updated in March 2023, should be 
    revised based on the issues discussed in Section~\ref{sec:datadictionary}.

    \item The precision of latitude, longitude, and location fields 
    should be reduced. We recommend limiting precision to four 
    decimal places (0.000X), which represents approximately 36.4 feet. 
    More decimal points strain credibility.

    \item The 11 address-related fields (\texttt{incident\_address}, 
    \texttt{street\_name}, \texttt{city}, \texttt{intersection\_street\_1 \& 2}, 
    \texttt{cross\_street\_1 \& 2}, \texttt{landmark}, \texttt{road\_ramp}, 
    \texttt{bridge\_highway\_segment}, and \texttt{taxi\_pick\_up\_location}) 
    should be standardized to USPS format. Software can easily 
    perform this standardization.

    \item There are multiple ways missing data is represented, such as 
    ``UNKNOWN'', ``N/A'', ``NA'', ``na'', and blanks (`` ''). Protocols 
    should be established to standardize the representation of missing data.
\end{itemize}

@@

\bibliographystyle{jds}
\bibliography{ref}

>>>>

\section{NYC 311 SR Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
\jy{should we mention the system upgrade just before the pandemic to
  document this historical event?}
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system.

@@

\section{NYC 311 SR Data Cleansing Issues} 
\label{sec:issues}

@@

\section{Government Open Data Curation Recommendations}
\label{sec:recommendations}
Building upon the insights gained from our case study of the 
NYC 311 Service Request (SR) data, we propose a set of data 
curation principles tailored for government-released open datasets. 
These principles are designed to address the unique challenges 
and requirements observed in the curation of such datasets, 
ensuring that they remain reliable, consistent, and useful for 
public services, research, and other applications.

\subsection{Principles}
\paragraph{Principle 1: Ensuring Consistency Across Agencies and Time}
A key challenge in government-released datasets is the 
harmonization of data fields across various agencies and 
maintaining this consistency over time. As seen in the NYC 
311 data, discrepancies in field names, formats, and definitions, 
such as those between \texttt{police\_precinct} and 
\texttt{police\_precincts}, can introduce significant 
inconsistencies. To address this, agencies should adopt 
standardized naming conventions and formats across all datasets. 
Regular audits and data harmonization processes should be 
implemented to ensure consistency across long-term datasets, 
particularly as agency structures or reporting standards evolve. 
For instance, ensuring that historical datasets reflect the same 
field naming conventions over a span of years can prevent 
inconsistent analyses.

\paragraph{Principle 2: Guaranteeing Accuracy and Validity in Data Entry}
Data accuracy and validity are fundamental to the utility of 
open datasets. In the NYC 311 data, issues such as 
invalid zip codes (e.g., 58\% invalid \texttt{zip\_codes} entries) and 
nonsensical date entries (e.g., negative durations) highlight the need 
for rigorous validation processes. Governments should establish 
clear protocols for data entry, ensuring that fields are populated with 
valid, legally acceptable values. For example, fields like 
\texttt{created\_date} and \texttt{closed\_date} should be subject to 
automated checks that prevent logical errors, such as closing an 
SR before it was opened. Ensuring the integrity of critical fields 
like geographic locations and agency names enhances the quality 
and reliability of the data.

\paragraph{Principle 3: Optimizing Storage and Representation Efficiency}
Efficient storage and data representation are crucial for handling 
large government datasets. By eliminating redundant fields and 
encoding categorical variables, agencies can significantly reduce 
file sizes and improve the efficiency of data analysis. In the NYC 
311 dataset, for example, redundant fields such as \texttt{borough} 
and \texttt{park\_borough} could be consolidated without any data 
loss, saving space and reducing file size by up to 12.3\%. Another 
example is the \texttt{taxi\_company\_borough} field, which is 99.94\% 
blank and should be evaluated to determine if its inclusion is 
necessary. Similarly, categorical variables like \texttt{complaint\_type} 
could be encoded in a standardized format to optimize storage and 
streamline analyses. Reducing file size not only saves storage space 
but also enhances the performance of queries and analysis processes.

\paragraph{Principle 4: Updating Data Dictionaries and Documentation}
Clear and accurate documentation is essential for the effective use 
of government-released datasets. The NYC 311 data revealed several 
issues where the Data Dictionary did not accurately reflect the 
actual data, such as missing field descriptions for computed fields. 
Updating the Data Dictionary to include descriptions of all fields, 
especially computed fields like \texttt{@computed\_region\_xxxx}, 
would improve transparency and usability. Regular updates should 
be made to ensure that any changes in data structure or content are 
reflected. This practice enhances the ability of external users, 
including researchers and developers, to understand and work 
effectively with the data.

\paragraph{Principle 5: Automating Quality Assurance Processes}
Automating data validation and quality assurance processes can 
significantly improve the reliability of government datasets. In the 
NYC 311 case, issues such as large spikes in SR creation and closure 
at exactly midnight or noon likely stem from automated processes that 
distort the accurate capture of SR durations. By implementing 
real-time validation tools, governments can prevent such anomalies 
from entering the dataset. Automated systems can be designed to 
flag inconsistencies, incorrect values, or illogical sequences, such 
as invalid \texttt{incident\_zip} values or illogical date entries. 
Automation can also facilitate faster detection and correction of 
errors, reducing the need for manual intervention and ensuring 
the data remains clean and reliable as it is continuously updated.

\paragraph{Principle 6: Addressing Data Transparency and Accessibility}
Transparency is a core tenet of open data, and governments must ensure 
that datasets are not only accurate but also accessible and easily 
understandable. This involves providing clear metadata and making 
datasets available in user-friendly formats. For example, in the NYC 
311 dataset, fields like \texttt{@computed\_region\_xxxx\_xxxx} and 
the distinction between \texttt{incident\_zip} and \texttt{zip\_codes} 
can be confusing for non-expert users. These fields should be clearly 
documented, explaining how they are calculated and how they differ. 
Additionally, governments should communicate transparently about 
data quality issues, such as notifying the public when errors in SR 
fields (e.g., erroneous closed dates) are discovered and explaining 
how such issues will be resolved. By ensuring transparency and 
keeping the public informed, governments foster trust and promote 
wider use of datasets for research, civic engagement, and innovation.

\subsection{Actionable Steps}
While these principles are broad, actionable steps are required to 
implement them effectively. Governments can begin by investing in 
real-time validation systems, regular updates to data dictionaries, 
and tools for encoding and standardizing data fields. For example, 
systems that automate the enforcement of data validation rules 
(e.g., flagging incorrect zip codes) can significantly improve data 
quality. Collaboration with city agencies and stakeholders is crucial 
to ensure that data standards, formats, and metadata meet the needs 
of diverse users, including policymakers, researchers, and the 
general public. Regular communication between agencies and external 
users can help identify recurring issues and areas where improvements 
are needed. By facilitating regular communication and collaboration 
with various stakeholders, governments can ensure data remains relevant 
and actionable for public use.

Engagement with data scientists and statisticians is key to improving 
data curation strategies. These experts can help design and refine 
curation processes, such as creating algorithms for automatic data 
cleaning, identifying inconsistent data entries, and ensuring the 
dataset’s structure is maintained over time. Their expertise is also 
critical in developing protocols for validating incoming data, correcting 
errors in real-time, and ensuring consistency across multiple data sources. 
Feedback from statisticians and data scientists, such as the issues reported 
in this paper, needs a clear and accessible channel to reach the 311 SR data 
team and other relevant city agencies. Establishing such feedback loops 
ensures that recommendations for improving data curation and validation 
are not only heard but implemented. This feedback mechanism would also allow 
ongoing dialogue between data users and city officials, leading to more dynamic 
and responsive data management practices.

A particularly valuable avenue for engagement is through events such as NYC 
Open Data Week, which fosters collaboration and innovation across sectors. 
Our involvement in this project stemmed from such an event, where we were 
inspired by the potential of open data to enhance public services. Governments 
should continue supporting these kinds of initiatives, which not only promote 
data literacy but also encourage community-driven improvements to datasets. 
By leveraging the expertise of data scientists, statisticians, stakeholders, 
and civic organizations, governments can ensure that their open datasets 
remain accurate, transparent, and useful for a wide range of applications.

% \begin{itemize}
% \item  Undertake a dedicated effort 
%     to investigate these data anomalies, confirm the veracity of the 
%     issues, and collaborate with the appropriate Agency Open Data 
%     Representatives to find resolutions. This effort may be complex 
%     and require software changes. Coordination and mutual approval 
%     of changes will be key, requiring strong support from senior 
%     leadership.

%     \item Standards and protocols should be established, agreed upon, 
%     and enforced. This includes defining the domain of legal values, 
%     ensuring data accuracy, implementing address normalization, 
%     clarifying data usage, and providing detailed data definitions.

%     \item Resolve many potential data integrity issues
%     through enforcement by software. For instance, software should 
%     prevent invalid entries like erroneous zip codes. Similarly, fields 
%     like \texttt{created\_date} and \texttt{closed\_date}, which 
%     measure responsiveness, should have logic checks to prevent 
%     illogical sequences, such as an SR being ``closed'' before 
%     being ``opened.''

%     \item The large spikes in SR creation and closure at midnight and 
%     noon likely point to automated processes from external Agency 
%     integrations. These distortions should be eliminated in favor of 
%     capturing accurate SR durations with precise timestamps.

%     \item As discussed, there are six ``computed'' fields that are not 
%     listed in the Data Dictionary. Some of these fields contain 
%     significant inaccuracies. We recommend updating the Data 
%     Dictionary and, if necessary, masking these fields from public use 
%     until their accuracy is improved.

%     \item Duplicate fields should be reviewed for removal. In some 
%     cases, such as \texttt{borough} and \texttt{park\_borough}, the 
%     duplication is 100\%. In others, the match is 80-90\%. 
%     These near-duplicates should be assessed to determine their 
%     relevance in the 311 SR dataset. Initial analysis suggests that 
%     removing these fields could reduce file size by up to 12\%, with 
%     further reductions possible upon deeper review.

%     \item Fields with very low data presence, such as 
%     \texttt{taxi\_company\_borough}, populated only 0.06\% of the time, 
%     should be evaluated to determine if they are necessary.

%     \item The Data Dictionary, last updated in March 2023, should be 
%     revised based on the issues discussed in Section~\ref{sec:datadictionary}.

%     \item The precision of latitude, longitude, and location fields 
%     should be reduced. We recommend limiting precision to four 
%     decimal places (0.000X), which represents approximately 36.4 feet. 
%     More decimal points strain credibility.

%     \item The 11 address-related fields (\texttt{incident\_address}, 
%     \texttt{street\_name}, \texttt{city}, \texttt{intersection\_street\_1 \& 2}, 
%     \texttt{cross\_street\_1 \& 2}, \texttt{landmark}, \texttt{road\_ramp}, 
%     \texttt{bridge\_highway\_segment}, and \texttt{taxi\_pick\_up\_location}) 
%     should be standardized to USPS format. Software can easily 
%     perform this standardization.

%     \item There are multiple ways missing data is represented, such as 
%     ``UNKNOWN'', ``N/A'', ``NA'', ``na'', and blanks (`` ''). Protocols 
%     should be established to standardize the representation of missing data.
% \end{itemize}

@@

\bibliographystyle{jds}
\bibliography{refs}
